\nopagenumbers \font\title=cmti12
\def\ve{\vfill\eject}
\def\vv{\vfill}
\def\vs{\vskip-2cm}
\def\vss{\vskip10cm}
\def\vst{\vskip13.3cm}

% \def\ve{\bigskip\bigskip}
% \def\vv{\bigskip\bigskip}
% \def\vs{}
% \def\vss{}
% \def\vst{\bigskip\bigskip}

\hsize=19cm
\vsize=27.58cm
\hoffset=-1.6cm
\voffset=0.5cm
\parskip=-.1cm
\ \vs \hskip -6mm TN410 AA15/16\ (Elementary Number Theory)\hfill Session C \hfill Rome, September 16, 2016 \hrule
\bigskip\noindent
{\title Family Name}\  \dotfill\ {\title Name}\ \dotfill {\title
Student ID(Matricola):}\ \dotfill\
\smallskip  \noindent
Solve the problems adding to the replies short and essential explenations. 
\it Please write the solutions in the designed areas. NO EXTRA SHEETS WILL BE ACCEPTED. 
\rm 1 Problem = 4 marks. Duration: 2 hours. No questions allowed in the first hour and in 
the last 20 minutes.
\smallskip
\hrule\smallskip
\centerline{\hskip 6pt\vbox{\tabskip=0pt \offinterlineskip
\def \trl{\noalign{\hrule}}
\halign to371.7pt{
\strut#& \vrule#\tabskip=1.1em plus 2.9em& \hfil#
       & \vrule#& \hfill#\hfil
       & \vrule#& \hfil#
       & \vrule#& \hfill#\hfil
       & \vrule#& \hfil#
       & \vrule#& \hfill#\hfil
       & \vrule#& \hfil#
       & \vrule#& \hfill#\hfil
       & \vrule#& \hfil#
       & \vrule#& \hfill#\hfil
       & \vrule#&\hfil#
       & \vrule#&\hfil#
       & \vrule#& \hfil#
       &\vrule#\tabskip=0pt\cr
\trl && 1 && 2 && 3a && 3b && 4 && 5 && 6 && 7 && 8 && TOTAL&\cr
\trl && &&   &&     &&   &&   &&  &&   &&    && && &\cr 
&&   &&   &&   &&     &&   && && && && && &\cr
\trl }}}
\medskip


\item{1.} Determine the number of solutions modulo $3^3\times 5^3$of the congruence $X^{20}-1\equiv0(\bmod 3^2\times 5^2)$ .\vv

\item{2.} Prove that a real number admits a finite continued fraction if and only if it is rational.\ve\vs

%\item{3.} \vv

\item{3.}  Let $N$ be a fixed positive integer and consider the aritmetic function $f_N(n):=\gcd(n,N)$.
\medskip
\itemitem{a.}  Show that $f_N(n)$ is a multiplicative function. When is it completely multiplicative?\vv

\itemitem{b.} Compute an asymptotic formula with an error term for $\sum_{n\le T}f_N(n)$.\vv

\item{4.} Show that if $N>4$ is not of the form $p^\alpha\ (p>3)$ or of the form $2p^\alpha\ (p>3)$, then there does not exists a primitive root modulo $N$. \ve\vs

\item{5.} Determine all primitive roots modulo $2\times 19^4$,\vv

\item{6.} Determine all integers $n$ such that $\sigma(n)=16$.\vv

\item{7.} Determine a formula for the number of pairs in $(x,y)\in{\bf Z}^2$ such that $x^2+y^2=2^a5^c13^d$.\ve \vs

\item{8.} Give a complete proof of Lagrange Theorem that states that every integer can be written as the sum of at most four squares.
\ \vst\bye
