\nopagenumbers \font\title=cmti12
\def\ve{\vfill\eject}
\def\vv{\vfill}
\def\vs{\vskip-2cm}
\def\vss{\vskip10cm}
\def\vst{\vskip13.3cm}

%\def\ve{\bigskip\bigskip}
%\def\vv{\bigskip\bigskip}
%\def\vs{}
%\def\vss{}
%\def\vst{\bigskip\bigskip}

\hsize=19.5cm
\vsize=27.58cm
\hoffset=-1.6cm
\voffset=0.5cm
\parskip=-.1cm
\ \vs \hskip -6mm Selected Topics on Finite Fields\hfill FINAL EXAM \hfill Lahore, November 3$^{\rm rd}$, 2011. \hrule
\bigskip\noindent
{\title Family Name}\  \dotfill\ {\title Given Name}\ \dotfill {\title
ROLL NO.}\ \dotfill\
\smallskip  \noindent
Solve the maximum number of problems providing, in each case, clear and synthetic explanations. \it Insert your answers in the
appropriate slots. Please NO SEPARATE SHEETS.
\rm 1 Exercise = 4 points. Total time: 2 hours.
\smallskip
\hrule\smallskip
\centerline{\hskip 6pt\vbox{\tabskip=0pt \offinterlineskip
\def \trl{\noalign{\hrule}}
\halign to277pt{\strut#& \vrule#\tabskip=0.7em plus 1em& \hfil#&
\vrule#& \hfill#\hfil& \vrule#& \hfil#& \vrule#& \hfill#\hfil&
\vrule#& \hfil#& \vrule#& \hfill#\hfil& \vrule#& \hfil#& \vrule#&
\hfill#\hfil& \vrule#& \hfil#& \vrule#& \hfill#\hfil& \vrule#&
\hfil#& \vrule#& \hfill#\hfil& \vrule#& \hfil#& \vrule#& \hfil#&
\vrule#\tabskip=0pt\cr\trl && SIGNATURE && 1 && 2 && 3 && 4 &&
5 && 6 && 7 && 8  &&  TOT. &\cr\trl && &&   &&
&&     &&   &&     &&   &&   &&    && &\cr &&
\dotfill &&       &&   &&   &&     &&   && && && &&
&\cr\trl }}}
\medskip

\item{1.} Answer to the following questions providing a justification of 1 line:\bigskip\bigskip\bigskip


\itemitem{a.} Is it true that there exists a unique field, up to isomorphisms, with 26 elements?\medskip\bigskip\bigskip

\ \dotfill\ \bigskip\bigskip\bigskip\vfil

\itemitem{b.} Is it true that if $f,g\in{\bf F}_q[X]$ and $(f-g)$ is divisible by $X^q-X$, then, as functions from ${\bf F}_q$
to ${\bf F}_q$,
$f$ and $g$ are the same?\medskip\bigskip\bigskip

\ \dotfill\ \bigskip\bigskip\bigskip\vfil

\itemitem{c.} Give an example of an irreducible polynomial of degree $2$ in ${\bf F}_5$.\medskip\bigskip\bigskip
 
\ \dotfill\ \bigskip\bigskip\bigskip\vfil

\itemitem{d.} When does a cyclotomic polynomial split in ${\bf F}_q[X]$?.\medskip\bigskip\bigskip

\ \dotfill\ \bigskip\bigskip\bigskip


\vfil\eject

%Dimostrare che un estensione finita \`{e} necessariamente algebrica. Produrre
%un esempio di un estensione algebrica non finita.

\item{2.} After having given the definition of permutation polynomial in ${\bf F}_q[X]$, show that
all linear polynomials are permutation polynomials.\vv


\item{3.} Describe all permutation polynomials in ${\bf F}_3[X]$ and ${\bf F}_4[X]$. 

\ve\ \vs

%Dopo aver verificato che \`e algebrico, calcolare
%il polinomio minimo di $\cos \pi/9$ su ${\bf Q}$.

\item{4.}  Let $N(a,q)$ be the number of irreducible polynomials of degree $a$ with coefficients in ${\bf F}_q$. 
Prove that if $a$ is prime, then $N(a,q)={q^a-q\over a}$.\vv

\item{5.} Let ${\bf F}={\bf F}_2[\alpha], \alpha^3=\alpha+1$. After having proven that ${\bf F}$
is a field, having computed its characteristic and its cardinality, compute the multiplicative order of $\alpha+1$.

\ve\ \vs

\item{6.} Consider the equation $2x^3+3x^2y+5zt^2-10=0$. Prove that it has at least $7$ solutions $(x_0,y_0,z_0,t_0)\in{\bf F}_7^4.$
{\it hint: This is an application of the Chevalley's Theorem.}
\vv\vv


\item{7.}  State a Theorem that you judge of great importance about finite fields.
\vv\vv

\item{8.} What is the numer of solutions of the equation $x^2+2y^2+3z^2=3$ over ${\bf F}_7$? 

\vv



\ \vst
 \bye
