\nopagenumbers \font\title=cmti12
\def\ve{\vfill\eject}
\def\vv{\vfill}
\def\vs{\vskip-2cm}
\def\vss{\vskip10cm}
\def\vst{\vskip13.3cm}

% \def\ve{\bigskip\bigskip}
% \def\vv{\bigskip\bigskip}
% \def\vs{}
% \def\vss{}
% \def\vst{\bigskip\bigskip}

\hsize=19.5cm
\vsize=27.58cm
\hoffset=-1.6cm
\voffset=0.5cm
\parskip=-.1cm
\ \vs \hskip -6mm CR410 AA14/15 (Public Key Crypto)\hfill Final Exam \hfill Rome, May 27, 2015. \hrule
\bigskip\noindent
{\title Family Name}\  \dotfill\ {\title Name}\ \dotfill {\title
Student ID(Matricola):}\ \dotfill\
\smallskip  \noindent
Solve the problems adding to the replies short and essential explenations. 
\it Please write the solutions in the designed areas. NO EXTRA SHEETS WILL BE ACCEPTED. 
\rm 1 Problem = 4 marks. Duration: 2 hours. No questions allowed in the first hour and in 
the last 20 minutes.
\smallskip
\hrule\smallskip
\centerline{\hskip 6pt\vbox{\tabskip=0pt \offinterlineskip
\def \trl{\noalign{\hrule}}
\halign to248pt{\strut#& \vrule#\tabskip=0.7em plus 1em& \hfil#&
\vrule#& \hfill#\hfil& \vrule#& \hfil#& \vrule#& \hfill#\hfil&
\vrule#& \hfil#& \vrule#& \hfill#\hfil& \vrule#& \hfil#& \vrule#&
\hfill#\hfil& \vrule#& \hfil#& \vrule#& \hfill#\hfil& \vrule#&
\hfil#& \vrule#& \hfill#\hfil& \vrule#& \hfil#& \vrule#& \hfil#&
\vrule#\tabskip=0pt\cr\trl && 1 && 2 && 3 && 4 &&
5 && 6 && 7 && 8  && 9 && TOT. &\cr\trl  &&   &&
&&  &&   &&   &&     &&  &&    &&  && &\cr &&       &&   &&      &&   && && && &&
 && && &\cr\trl }}}
\medskip

\item{1.} Answer to the following questions with a justification of one line:\bigskip
\itemitem{a.} Is it true that if $n$ is odd, the irreducible factors of $x^{3^{n}}-x$ have all odd degree?\bigskip

\ \dotfill\ \vfil\bigskip\bigskip

\itemitem{b.} Is it true that it is not possible to elliptic curves over finite fields for the Diffie -- Hellman
key exchange protocol?\bigskip

\ \dotfill\ \vfil\bigskip\bigskip

\itemitem{c.} Which is the probability that an irreducible polynomial of degree 5 over ${\bf F}_5$ is primitive?\bigskip
 
\ \dotfill\ \vfil\bigskip\bigskip

\itemitem{d.} Is it true that there are no elliptic curves over ${\bf F}_{{11}^2}$ such that 
$E({\bf F}_{11^2})\cong {\bf Z}/7{\bf Z}\oplus{\bf Z}/42{\bf Z}$? \bigskip

\ \dotfill\ \bigskip\bigskip\bigskip

\item{2.} After having recalled the RSA cryptosystem, decode the crypted text ${\cal C}=25$ 
knowing tha the public key is $(7,143)$. 
\vfill\eject\ \vskip-2cm

\item{3.} Expain how the Miller -- Rabin primality test works.\vfill

\item{4.} Explain the Goldwasser -- Micali cryptosystem.\vfill\eject\ \vskip-2cm


\item{5.} After having defined Carmichael numbers and having recalled their main properties, prove that 
$6601$ is Carmichael.
\vfill

\item{6.} Let $E: y^2+\alpha y=x^3$ be and elliptic curve over ${\bf F}_8={\bf F}_2[\alpha], \alpha^3=\alpha+1$. Determine
$\#E({\bf F}_2[\alpha])$ and $\#E({\bf F}_{2^6})$. What can be said regarding the group structure?
\vfill\eject\ \vskip-2cm

\item{7.} Determine the structure of group of rational points of an elliptic curve  $E$ over ${\bf F}_{53^2}$ knowing that
it contains a point of order $392$ and a unique point of order $2$.\vfill\vfill

\item{8.} Consider the elliptic curve $E: y^2=x^3+x+1$ and compute $\#E({\bf F}_{3^{100}})$ .\vfill\vfill

\item{9.} Prove that an elliptic curve defined on a field of chaacteristic  $3$ has at most two points of order $3$.\hfill\break 
%{\scriptsize
$\ {}\qquad\qquad$\hfill{\bf hint:} \it Study the identity 
$2P=-P$ for the Weierstrass equation $y^2=x^3+ax^2+bx+c$.
%}
\vfil\ \vst\bye
