\nopagenumbers \font\title=cmti12
\def\ve{\vfill\eject}
\def\vv{\vfill}
\def\vs{\vskip-2cm}
\def\vss{\vskip10cm}
\def\vst{\vskip13.3cm}

% \def\ve{\bigskip\bigskip}
% \def\vv{\bigskip\bigskip}
% \def\vs{}
% \def\vss{}
% \def\vst{\bigskip\bigskip}

\hsize=19.5cm
\vsize=27.58cm
\hoffset=-1.6cm
\voffset=0.5cm
\parskip=-.1cm
\ \vs \hskip -6mm CR410 AA14/15 (Public Key Crypto)\hfill Midterm Exam \hfill Rome, March 31, 2015. \hrule
\bigskip\noindent
{\title Family Name}\  \dotfill\ {\title Name}\ \dotfill {\title
Student ID(Matricola):}\ \dotfill\
\smallskip  \noindent
Solve the problems adding to the replies short and essential explenations. 
\it Please write the solutions in the designed areas. NO EXTRA SHEETS WILL BE ACCEPTED. 
\rm 1 Problem = 4 marks. Duration: 2 hours. No questions allowed in the first hour and in 
the last 20 minutes.
\smallskip
\hrule\smallskip
\centerline{\hskip 6pt\vbox{\tabskip=0pt \offinterlineskip
\def \trl{\noalign{\hrule}}
\halign to320.5pt{
\strut#& \vrule#\tabskip=1.1em plus 2.6em& \hfil#
       & \vrule#& \hfill#\hfil
       & \vrule#& \hfil#
       & \vrule#& \hfill#\hfil
       & \vrule#& \hfil#
       & \vrule#& \hfill#\hfil
       & \vrule#& \hfil#
       & \vrule#& \hfill#\hfil
       & \vrule#& \hfil#
       & \vrule#& \hfill#\hfil
       & \vrule#&\hfil#
       & \vrule#& \hfil#
       &\vrule#\tabskip=0pt\cr
\trl && 1 && 2 && 3 && 4 && 5 && 6 && 7 && 8 && TOTAL&\cr
\trl && &&   &&     &&   &&   &&  &&   &&    && &\cr 
&&   &&   &&   &&     &&   && && && && &\cr
\trl }}}
\medskip


\item{1.} Answer the following questions providing a justification of one line:\bigskip
\itemitem{a.} Is the Diffie-Hellmann key exchange protocol only defined for the cyclic group ${\bf F}_{p^n}^*$?\medskip\bigskip\bigskip

\ \dotfill\ \vfil

\itemitem{b.} Is it true that there are non-isomorphic finite fields in which the respective multiplicative groups
have the same number of generators?\medskip\bigskip\bigskip

\ \dotfill\ \vfil

\itemitem{c.} If $f,g\in{\bf F}_p[x]$ have the same degree, is it true that the splitting field of $f$
contains the roots of $g$?\medskip\bigskip\bigskip
 
\ \dotfill\ \vfil

\itemitem{d.} Write down all irreducible polinomials in ${\bf F}_2[x]$ with degree less or equal to $4$.\medskip\bigskip\bigskip

\ \dotfill\ \bigskip\bigskip\bigskip

\item{2.} After having written recursive formulas for the calculation of the Bezout identity between two integers, compute
that identity for $(1345,9875)$.  Next compute the greatest common divisor $(1345.9875)$ using the binary algorithm.\ve\vs

\item{3.} After having shown that $3$ is a primitive root modulo $31$, campute the discrete logarithm 
 $\log_32\in{\bf Z}/30{\bf Z}$ using the algorithm Baby Steps Giant Steps.\vv

\item{4.} Outline some cryptographic systems that base their security on the problem of
discrete logarithm .\ve\vs

\item{5.} Determine all integers $X$ in the interval  $[-200,10]$ such that $\cases{X\equiv 2\bmod 4\cr X\equiv 4\bmod 5\cr 3X\equiv 4\bmod 7.}$ \vv

\item{6.}  Provide an example of explicit finite field for $32$ elements and among its elements determine a primitive root.\ve \vs

\item{7.} Determine the degree over ${\bf F}_{13}$ of the splitting field of the polynomial
$$(T^{13^8}-27T^{13^5}+26T^{13^{4}})(T^2+13T+27)(T^3+14)(T^{13^8}+25T^{13})\in{\bf F}_{13}[T].$$\vv

\item{8.} After having explained briefly the algorithm of successive squares, compute $\alpha^{1047}\in{\bf F}_7[\alpha], \alpha^3=\alpha-2$.
 
\ \vst\bye
