\documentclass[12pt]{amsart}
\usepackage{amssymb}
%\textwidth=7in \oddsidemargin=-1.1cm \evensidemargin=-0.9cm
% ----------------------------------------------------------------
\begin{document}


\title[Reviewed Pubblications]{List of (to be) Reviewed Pubblications}%
\author{Francesco Pappalardi}
\address{Dipartimento di Matematica, Universit\`{a} Roma Tre,
Largo S. L. Murialdo, 1, I--00191, Roma ITALIA}
\email{pappa@mat.uniroma3.it}

\thanks{Updated \today}%
\subjclass{11Xxx}%
\keywords{Number Theory}%

\date{\today}%
%\begin{abstract}
%\end{abstract}
\vspace*{-2cm} \maketitle \thispagestyle{empty} \nocite{*}
\vspace*{-.5cm}
% ----------------------------------------------------------------
\bibliographystyle{unsrt}
\begin{thebibliography}{10}

\bibitem{0} \textbf{F\!P} and Alf van der Poorten:
\textit{Pseudo-Elliptic Integrals, Units and Torsion}.
Preliminary.

\bibitem{1} \textbf{F\!P}: \textit{A survey on $k$--power freeness}.
To appear in Proceeding of the Conference in Analytic Number
Theory in Honor of Prof. Subbarao at I.M.Sc. Chennai, January
2003.\bigskip

\bibitem{2} W. Banks and \textbf{F\!P}: \textit{Powerfree Values of the Euler Function.}
\newblock Submitted.\bigskip

\bibitem{3} F. Luca  and \textbf{F\!P}: \textit{Members of binary recurrences on lines of the Pascal
triangle}.  \newblock Submitted.\bigskip

\bibitem{4} S. D. Adhikari, J. B. Friedlander, S. Konyagin  and \textbf{F\!P}:
\textit{Contributions to zero-sum problems}. \newblock
Submitted.\bigskip

\bibitem{5} S. Konyagin and \textbf{F\!P}: \textit{Enumerating permutation polynomials over finite
fields by degree II.} \newblock Submitted.\bigskip

\bibitem{6} \textbf{F\!P}: \textit{Squarefree Values of the Order Function}.
\newblock {New York Journal of Mathematics} \textbf{9} (2003), 331-344.\bigskip

\bibitem{7} C. David and \textbf{F\!P}: \textit{Average Frobenius Distribution for inerts in
$\mathbb Q(i)$}. \newblock Submitted.\bigskip

\bibitem{daniele} A. Cosentino and \textbf{F\!P}: \textit{Fast
computation of $k$--dimensional residues and an application to
probabilistic encryption}. \newblock Submitted.\bigskip

\bibitem{igorn} \textbf{F\!P}: Filip Saidak and Igor E. Shparlinski:
\textit{Squarefree Values of the Carmichael Function}.
\newblock {J. Number Theory} \textbf{103}-1 (2003) 122--131.\bigskip

\bibitem{Claudia2} Claudia Malvenuto and \textbf{F\!P}:
\textit{Enumerating Permutation Polynomials II: $k$--cycles with
minimal degree}. \newblock To appear in {Finite Fields
Appl.}\bigskip

\bibitem{Claudia1} Claudia Malvenuto and \textbf{F\!P}:
\textit{Enumerating Permutation Polynomials I: Permutations with
Non-Maximal Degree}. \newblock {Finite Fields Appl.} \textbf{8}
(2002) 531--547.\bigskip

\bibitem{Sergey} Sergei Konyagin and \textbf{F\!P}: \textit{Enumerating
Permutation Polynomials over Finite Fields by Degree.} \newblock
{Finite Fields Appl.} \textbf{8} (2002) 548--553.\bigskip

\bibitem{Sukumar} Sukumar Das Adhikari and \textbf{F\!P}: \textit{On the
visibility problem in the function field case.} \newblock {C. R.
Math. Acad. Sci. Soc. R. Can.} \textbf{24} (2002), no. 3,
109--116.\bigskip

\bibitem{Mans-On-graphs} Bernard~Mans, \textbf{F\!P} and Igor
Shparlinski. \textit{On the spectral \'{A}d\'{a}m property for
circulant graphs.} \newblock {Discrete Mathematics} \textbf{254}
(2002) 309--329.\bigskip

\bibitem{MR2001f:11052} Ernest~S. Croot, III, David~E. Dobbs, John~B.
Friedlander, Andrew~J. Hetzel and \textbf{F\!P}: \textit{Binary
{E}gyptian fractions}. \newblock {J. Number Theory} \textbf{84}-1
(2000) 63--79.\bigskip

\bibitem{MR2000i:11149} Leonardo Cangelmi and \textbf{F\!P}: \textit{On
the $r$-rank {A}rtin conjecture. {I}{I}.} \newblock {J. Number
Theory} \textbf{75}-1 (1999) 120--132.\bigskip

\bibitem{MR2000h:11056} Chantal David, Hershy Kisilevsky, and
\textbf{F\!P}: \textit{Galois representations with non-surjective
traces.} \newblock {Canad. J. Math.}, 51(5):936--951,
1999.\bigskip

\bibitem{MR2000g:11045} Chantal David and \textbf{F\!P}: \textit{Average
{F}robenius distributions of elliptic curves.} \newblock
{Internat. Math. Res. Notices} \textbf{4} (1999) 165--183.\bigskip

\bibitem{MR98c:11102} \textbf{F\!P}: \textit{On {H}ooley's theorem with
weights.} {Rend. Sem. Mat. Univ. Politec. Torino}, \textbf{53}-4
(1995) 375--388 in \newblock \textit{Number theory, II} (Rome,
1995).\bigskip

\bibitem{MR97m:11003} \textbf{F\!P}, Alberto Perelli, and Carlo Viola,
editors. \newblock {\em Number theory. {I}{I}}, Turin, 1995.
Libreria Editrice Universitaria Levrotto \& Bella. \newblock Rend.
Sem. Mat. Univ. Politec. Torino {\bf 53} (1995), no. 4.\bigskip

\bibitem{MR97m:11002} \textbf{F\!P}, Alberto Perelli, and Carlo Viola,
editors. \newblock {\em Number theory. {I}}, Turin, 1995. Libreria
Editrice Universitaria Levrotto \& Bella. \newblock Rend. Sem.
Mat. Univ. Politec. Torino {\bf 53} (1995), no. 3.\bigskip

\bibitem{MR97g:11132} \textbf{F\!P} and Igor Shparlinski. \textit{On
{A}rtin's conjecture over function fields.} \newblock {Finite
Fields Appl.} \textbf{1}-4 (1995) 399--404.\bigskip

\bibitem{MR97f:11082} \textbf{F\!P}: \textit{On the $r$-rank {A}rtin
conjecture.} \newblock {Math. Comp.} \text66-218 (1997)
853--868.\bigskip

\bibitem{MR97d:11141} \textbf{F\!P}: \textit{On the order of finitely
generated subgroups of ${\mathbf{Q}}\sp
 *\pmod p$ and divisors of $p-1$.} \newblock {J. Number Theory}
\textbf{57}-2 (1996) 207--222.\bigskip

\bibitem{MR96k:11120} \textbf{F\!P}: \textit{On minimal sets of
generators for primitive roots.} \newblock {Canad. Math. Bull.}
\textbf{38}-4 (1995) 465--468.\bigskip

\bibitem{MR96j:11158} Hershy Kisilevsky and \textbf{F\!P}: \textit{On
the exponent of the ideal class groups of imaginary extensions of
  $\mathbf {F}\sb q(x)$.} \newblock {Acta Arith.} \textbf{72}-4 (1995)
311--321.\bigskip

\bibitem{MR95d:11154} \textbf{F\!P}: \textit{On the exponent of the
ideal class group of $\mathbb Q(\sqrt{-d})$.} \newblock {Proc.
Amer. Math. Soc.} \textbf{123}-3 (1995) 663--671.\bigskip

\bibitem{MR93d:13005} Marco Fontana and \textbf{F\!P}: \textit{The prime
spaces as spectral spaces.} \newblock {Ann. Mat. Pura Appl. (4)},
\textbf{160} (1992) 331--345.\bigskip

\bibitem{MR92d:55020} Heddy.~E.~A. Campbell, Ian~Hughes, \textbf{F\!P} and
Paul~S. Selick. \textit{On the ring of invariants of ${\mathbf
{F}}\sp *\sb {2\sp n}$.}\newblock {Comment. Math. Helv.}
\textbf{66}-2 (1991) 322--331.\bigskip

\bibitem{MR1683339} Bernard Mans, \textbf{F\!P}, and Igor Shparlinski.
\textit{On the \'{A}d\'am conjecture on circulant graphs.} \newblock In
{\em Computing and combinatorics (Taipei, 1998)}, pages 251--260.
  Springer, Berlin, 1998.\bigskip
\end{thebibliography} \end{document}
